---
title: "estonian-lakes"
author: "Justin Pomeranz"
date: "2023-10-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
est_dat <- read_csv(here::here("data/noges_noges_simple.csv"))

```

### Setup
```{r, eval=FALSE}
# estonian lakes from Noges and Noges 2014

library(tidyverse)

est_dat <- read_csv("data/noges_noges_simple.csv")

```

Custom function for hydro day
```{r}
# Custom function from Oleksy
# custom function to calculate julian-day of water year
hydro.day = function(x, start.month = 10L) {
  start.yr = year(x) - (month(x) < start.month)
  start.date = make_date(start.yr, start.month, 1L)
  as.integer(x - start.date + 1L)
}
```

Calc hydro days. Also compare their duration with the one we calculated
```{r}
est_dat <- est_dat %>%
  mutate(j_on_wy = hydro.day(Start),
         j_off_wy = hydro.day(End),
         year = year(End),
         duration_calc = j_off_wy - j_on_wy,
         dur_equal = Duration_orig == duration_calc) 

est_dat %>%
  filter(dur_equal == FALSE) %>%
  distinct(Lake) # just Mustvee calcs are different
```

One of the lakes has all the same durations, one does not. I use their original calculation in plot below

```{r}
est_dat %>%
  filter(dur_equal == FALSE) %>% 
  mutate(diff = duration_calc - Duration_orig) %>%
  summarize(mean(diff), sd(diff))
```
our calc versus theres is off by an average of 28 $\pm$ 19 days (mean $\pm$ 1 standard deviation)

# plots  
### Ice on by year  

```{r}
ggplot(est_dat,
       aes(x = year,
           y = j_on_wy,
           color = Lake)) +
  geom_point() +
  stat_smooth(method = "gam")
```

### Ice off by year  


```{r}
est_dat %>%
  ggplot(aes(x = year,
             y = j_off_wy,
             color = Lake)) +
  geom_point() +
  stat_smooth(method = "gam")
```


### Duration, original calc  

```{r}
est_dat %>%
  ggplot(aes(x = year,
             y = Duration_orig,
             color = Lake)) +
  geom_point() +
  stat_smooth(method = "gam")

```



### Duration, new calc  

```{r}
est_dat %>%
  ggplot(aes(x = year,
             y = duration_calc,
             color = Lake)) +
  geom_point() +
  stat_smooth(method = "gam")

```