---
title: "GAM ACF plots"
author: "Justin Pomeranz"
date: "October 2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# libraries
library(tidyverse)
```

# Auto-correlation for Generalized Additive Models for ice lake phenology
# Justin Pomeranz <jpomeranz@coloradomesa.edu>
# October 2023  

This html file contains ACF plots for 24 models (8 lakes * 3 response variables). 

I briefly looked into how to interpret these plots, and it *seems* ok, but I was hoping you could take a look at them as well since I think you know more about this than me. 

### libraries

```{r, eval=FALSE}
library(tidyverse)
```


### data

#### Note on the data 

This data is modified and wrangled in `gam_mods.rmd`. 

```{r}
full_data <- readRDS(file = here::here("scripts/all_lake_data.rds"))
```

### List of models 

These models were also created in `gam_mods.rmd`. 

```{r}
ice_response_list <- readRDS(
  file = here::here("scripts/ice_response_list.RDS"))
```


## Custom functions  

by Pomeranz  

### function to plot ACF
```{r}
plot_acf <- function(model, title, lake_name = NULL){
ACF <- acf(resid(model,
                   type = "response"),
             plot = FALSE)
ACF <- setNames(
  data.frame(
    unclass(ACF)[c("acf", "lag")]),
  c("ACF","Lag"))

plot <- ggplot(ACF, aes(x = Lag, y = ACF)) +
  geom_hline(aes(yintercept = 0)) +
  geom_segment(mapping = aes(xend = Lag, yend = 0)) +
  labs(title = title,
       subtitle = lake_name)

print(plot)
}
```

# function to scale up plotting  

```{r}
plot_acf_list <- function(ice_list, response_name){
  for(lake in 1:length(ice_list)){
  lake_name = names(ice_days_list)[lake]
  plot_acf(ice_list[[lake]]$mod, response_name, lake_name = lake_name)
}
}
```

# ACF plots

# ice duration  


```{r}
ice_days_list <- ice_response_list$duration
plot_acf_list(ice_days_list, "Duration")
```



# ice OFF  

```{r}
ice_off_list <- ice_response_list$off_date
plot_acf_list(ice_off_list, "Ice-off-date")
```


# ice ON  

```{r}
ice_on_list <- ice_response_list$on_date
plot_acf_list(ice_on_list, "ice-on-date")
```

