---
title: "GAM models"
author: "Justin Pomeranz"
date: "8/22/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# libraries
library(tidyverse)
library(lme4)
library(lmerTest)
library(ggeffects)
library(broom)
library(mgcv)
library(gratia)
library(broom)
```

# Generalized Additive Models for ice lake phenology
# Justin Pomeranz <jpomeranz@coloradomesa.edu>
# September 2023

### libraries

```{r}
library(tidyverse)
library(lme4)
library(lmerTest)
library(ggeffects)
library(broom)
library(mgcv)
library(gratia)
library(broom)
```

Custom function previously written by Bella Oleksy to calculate the number of days based on water year (beginning on October 1)

```{r}
# Custom function from Oleksy
# custom function to calculate julian-day of water year
hydro.day = function(x, start.month = 10L) {
  start.yr = year(x) - (month(x) < start.month)
  start.date = make_date(start.yr, start.month, 1L)
  as.integer(x - start.date + 1L)
}
```


### data

#### Note on the data 

I copied the raw data from: <https://github.com/afilazzola/IcePhenologyDatabase/blob/main/data/PhenologyData.csv#L6>

When I tried to download the original csv file, excel was doing weird things with the date columns. This occurred even when I copied/pasted directly into excel, and changed the advanced settings for "text to columns" option. 

I found a workaround:  
1. I copied the raw data from the github link above  
2. Pasted the raw values into notepad and saved as a `.txt` file  
3. read in the `.txt` file  
4. All dates correctly imported as `<date>` format.  

#### Wrangle data

* Select the columns  
* filter out appropriate lakes  
* calculate "hydro day" (Julian day of water year, starting on Oct 1)  
* filter years between 1927 to 2022  
* Calculate ice duration days  

Read in the lake phenology data from "other" lakes. 
```{r message=FALSE, warning=FALSE}
## Non-Yellowstone lakes
non_ysl <- read_csv(here::here("scripts/other_phenology.txt"))

non_ysl <- non_ysl %>%
  # select necessary columns
  select(lake, lakecode, start_year, iceOn, iceOff, orig_duration) %>% 
  # filter out lakes: 
  filter(lakecode == "JK25"| # Lake Haukivesi
           lakecode == "JK02"| # Lake Kallavesi
           lakecode == "GW369"| # Lake Kallsjön
           lakecode == "JK03"| # Näsijärvi
           lakecode == "JK05"| # Päijänne
           lakecode == "JK40"| # Pielinen
           lakecode == "NG1" # Lake Baikal
  ) %>%
  # caluclate on and off julian dates for water year
  mutate(j_on_wy = hydro.day(iceOn),
         j_off_wy = hydro.day(iceOff)) %>%
  # filter out data only from 1927 to 2022
  filter(start_year >= 1927, start_year <=2022) %>%
  mutate(ice_days = j_off_wy - j_on_wy)

```

#### Modify yellowstone lake data
```{r}
# yellowstone data
ysl_ice <- read.csv(here::here("scripts/YSLoff.csv")) %>%
  select(Year, IceOnDate, IceOffDate,) %>%
  mutate(lake = "yellowstone")

# modify data
ysl_ice <- ysl_ice %>%
  unite(IceOnDate, c(IceOnDate, Year), sep = "-", remove = FALSE) %>%
  unite(IceOffDate, c(IceOffDate, Year), sep = "-", remove = FALSE) %>%
  mutate(on_date = dmy(IceOnDate),
         off_date = dmy(IceOffDate)) %>%
  mutate(j_on_wy = hydro.day(on_date),
         j_off_wy = hydro.day(off_date),
         ice_days = j_off_wy - j_on_wy) %>%
  rename(start_year = Year) %>%
  select(lake, start_year, j_on_wy, j_off_wy, ice_days) 
```

#### combine the two data sets into `full_data`
```{r}
# combine data sets
full_data <- bind_rows(non_ysl, ysl_ice) %>%
  mutate(start_y_c = start_year - mean(start_year, na.rm = TRUE))
saveRDS(full_data,
        file = here::here("scripts/all_lake_data.rds"))
```


# GAMs

#### Function to fit GAMs

* Fit a GAM with specified `gam_formula`  

* Make the following objects  

  * model summaries  
  
  * draw_plot  
  
  * appraise model  
  
* Return a list with the following:  

  * `lake_name`  
  
  * `mod` = model fit  
  
  * `draw_plot` = results of the `draw()` function  
  
  * `appraise` = results of the `apraise()` function  
  

```{r}
gam_mod_fun <- function(data, gam_formula, lake_name){
  mod <- gam(gam_formula,
             family=Gamma(link="log"),
             data = data %>%
               filter(lake == lake_name),
             correlation = corCAR1(
               form = ~ start_year),
             method = "REML")
  mod_summary <- summary(mod)
  draw_plot <- draw(mod)
  app_plot <- appraise(mod)
  return(list(lake_name = lake_name,
              mod = mod,
              summary = mod_summary,
              draw_plot = draw_plot,
              appraise = app_plot))}
```

# Summary function  

This function summarizes the GAM model fits and displays output in a "tidy" format

```{r}
gam_summary <- function(mydata, gam_formula, group_var) {
  group_var = enquo(group_var)

  mydata <- mydata %>% 
    group_by(!!group_var) %>% 
    nest() 

  mydata %>% 
    mutate(
      model = map(data,
                  ~(gam(
                    gam_formula,
                    family=Gamma(link="log"),
                    data = .,
                    correlation = corCAR1(
                      form = ~ start_year),
                    method = "REML")))) 
}

```


# list of models

`for`-loop producing a list of the summary outputs for each lake. 

Run the loop for each response variable

### ice on
```{r}
lake_names <- full_data %>%
  pull(lake) %>%
  unique()

ice_on_list <- list()
for (lake in 1:length(lake_names)){
  lake_name = lake_names[lake]
  response_mod <- gam_mod_fun(full_data, j_on_wy~s(start_year), lake_name)
  ice_on_list[[lake]] <- response_mod
}
names(ice_on_list) <- lake_names
```

### ice off
```{r}
ice_off_list <- list()
for (lake in 1:length(lake_names)){
  lake_name = lake_names[lake]
  response_mod <- gam_mod_fun(full_data, j_off_wy~s(start_year), lake_name)
  ice_off_list[[lake]] <- response_mod
}
names(ice_off_list) <- lake_names
```


### duration
```{r}
ice_days_list <- list()
for (lake in 1:length(lake_names)){
  lake_name = lake_names[lake]
  response_mod <- gam_mod_fun(full_data, ice_days~s(start_year), lake_name)
  ice_days_list[[lake]] <- response_mod
}
names(ice_days_list) <- lake_names
```

## Save list of lists
ice_response_list = list(ice_days_list
ice_off_list
ice_on_list)

```{r}
ice_response_list = list(duration = ice_days_list,
                         off_date = ice_off_list,
                         on_date = ice_on_list)
saveRDS(ice_response_list, file = here::here("scripts/ice_response_list.RDS"))
```


# print out results of baikal as example

```{r}
list(ice_days_list$baikal,
     ice_off_list$baikal,
     ice_on_list$baikal)
```

## Yellowstone

```{r}
list(ice_days_list$yellowstone,
     ice_off_list$yellowstone,
     ice_on_list$yellowstone)
```
## plot results

**NOTE** these plots use the default `"gam"` option in `stat_smooth()`. They may differ slightly from the model fits used above. 

# Ice duration figure

```{r}
ggplot(full_data,
       aes( x = start_year,
            y = ice_days,
            color = lake)) +
         geom_point() +
         stat_smooth(method = "gam") +
  facet_wrap(~lake)
```

# Ice-on figure

```{r}
ggplot(full_data,
       aes( x = start_year,
            y = j_on_wy,
            color = lake)) +
         geom_point() +
         stat_smooth(method = "gam") +
  facet_wrap(~lake)
```

# Ice-off figure

```{r}
ggplot(full_data,
       aes( x = start_year,
            y = j_off_wy,
            color = lake)) +
         geom_point() +
         stat_smooth(method = "gam") +
  facet_wrap(~lake)
```

## summary output of gams

**NOTE** edf in the summary output indicates the "wiggly-ness" of the model.  
* edf 1 == Linear
* edf 2 == quadratic  
* edf 3 == cubic polynomial... 
* etc.  


Most of the following models fall between edf 1 to edf 3.  

Lake Baikal has edf 6 for ice-off. 


The following summary outputs are arranged by p-value. Lakes at the top are "more significant", and lakes lower down are less or non-significant. 

Yellowstone is non-significant for all three responses. Most lakes have a significant response for most response variables.   

### Ice-on
```{r}
gam_summary(full_data, j_on_wy~s(start_year), lake) %>%
  mutate(tidied = map(model, tidy)) %>%
  select(tidied) %>%
  unnest(tidied) %>%
  arrange(lake_name, p.value)
```

Kallsjon and Yellowstone are non-significant 



### Ice-off
```{r}
gam_summary(full_data, j_off_wy~s(start_year), lake) %>%
  mutate(tidied = map(model, tidy)) %>%
  select(tidied) %>%
  unnest(tidied) %>%
  arrange(lake_name, p.value)
```

Only Yellowstone is non-significant 

### duration
```{r}
gam_summary(full_data, ice_days~s(start_year), lake) %>%
  mutate(tidied = map(model, tidy)) %>%
  select(tidied) %>%
  unnest(tidied) %>%
  arrange(lake_name, p.value)
```

Kallsjon and Yellowstone are non-significant 


## auto corr

### Still need to scale this up to the full data

```{r}
ACF <- acf(resid(ice_on_list$baikal$mod, type = "response"), plot = FALSE)
ACF <- setNames(data.frame(unclass(ACF)[c("acf", "lag")]), c("ACF","Lag"))
ggplot(ACF, aes(x = Lag, y = ACF)) +
  geom_hline(aes(yintercept = 0)) +
  geom_segment(mapping = aes(xend = Lag, yend = 0))
```

```{r}
plot_acf <- function(model, title, lake_name = NULL){
ACF <- acf(resid(model,
                   type = "response"),
             plot = FALSE)
ACF <- setNames(
  data.frame(
    unclass(ACF)[c("acf", "lag")]),
  c("ACF","Lag"))

plot <- ggplot(ACF, aes(x = Lag, y = ACF)) +
  geom_hline(aes(yintercept = 0)) +
  geom_segment(mapping = aes(xend = Lag, yend = 0)) +
  labs(title = title,
       subtitle = lake_name)

print(plot)
}
```

```{r}
# plot_acf(ice_days_list[[1]]$mod,
#          "Ice-Duration",
#          "lake_name")

# for(lake in 1:length(ice_days_list)){
#   lake_name = names(ice_days_list)[lake]
#   plot_acf(ice_on_list[[lake]]$mod, "ice-duration", lake_name = lake_name)
# }

plot_acf_list <- function(ice_list, response_name){
  for(lake in 1:length(ice_list)){
  lake_name = names(ice_days_list)[lake]
  plot_acf(ice_on_list[[lake]]$mod, response_name, lake_name = lake_name)
}
}
```

## ACF plots

### ice duration

```{r}
plot_acf_list(ice_days_list, "Duration")
```

### ice OFF
```{r}
plot_acf_list(ice_off_list, "Ice-off-date")
```
### ice ON
```{r}
plot_acf_list(ice_on_list, "ice-on-date")
```

