---
title: "non YSL mixed effects"
author: "Justin Pomeranz"
date: "8/22/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# libraries
library(tidyverse)
library(lme4)
library(lmerTest)
library(ggeffects)
```

# mixed effects for ice lake phenology
# Justin Pomeranz <jpomeranz@coloradomesa.edu>
# August 2023

### libraries

```{r}
library(tidyverse)
library(lme4)
library(lmerTest)
library(ggeffects)
```

```{r}
# Custom function from Oleksy
# custom function to caluclate julian-day of water year
hydro.day = function(x, start.month = 10L) {
  start.yr = year(x) - (month(x) < start.month)
  start.date = make_date(start.yr, start.month, 1L)
  as.integer(x - start.date + 1L)
}
```


### data
```{r message=FALSE, warning=FALSE}
## Non-Yellowstone lakes
non_ysl <- read_csv(here::here("scripts/other_phenology.txt"))

non_ysl <- non_ysl %>%
  # select necessary columns
  select(lake, lakecode, start_year, iceOn, iceOff, orig_duration) %>% 
  # filter out lakes: 
  filter(lakecode == "JK25"| # Lake Haukivesi
           lakecode == "JK02"| # Lake Kallavesi
           lakecode == "GW369"| # Lake Kallsjön
           lakecode == "JK03"| # Näsijärvi
           lakecode == "JK05"| # Päijänne
           lakecode == "JK40"| # Pielinen
           lakecode == "NG1" # Lake Baikal
  ) %>%
  # caluclate on and off julian dates for water year
  mutate(j_on_wy = hydro.day(iceOn),
         j_off_wy = hydro.day(iceOff)) %>%
  # filter out data only from 1927 to 2022
  filter(start_year >= 1927, start_year <=2022) %>%
  mutate(ice_days = j_off_wy - j_on_wy)

# yellowstone data
ysl_ice <- read.csv(here::here("scripts/YSLoff.csv")) %>%
  select(Year, IceOnDate, IceOffDate,) %>%
  mutate(lake = "yellowstone")

# modify data
ysl_ice <- ysl_ice %>%
  unite(IceOnDate, c(IceOnDate, Year), sep = "-", remove = FALSE) %>%
  unite(IceOffDate, c(IceOffDate, Year), sep = "-", remove = FALSE) %>%
  mutate(on_date = dmy(IceOnDate),
         off_date = dmy(IceOffDate)) %>%
  mutate(j_on_wy = hydro.day(on_date),
         j_off_wy = hydro.day(off_date),
         ice_days = j_off_wy - j_on_wy) %>%
  rename(start_year = Year) %>%
  select(lake, start_year, j_on_wy, j_off_wy, ice_days) 

# combine data sets
full_data <- bind_rows(non_ysl, ysl_ice) %>%
  mutate(start_y_c = start_year - mean(start_year, na.rm = TRUE))
```


## models

* year has been centered and = fixed  

* lake = random  

Model Function. 

```{r}
mod_fun <- function(dat, response, title = NA){
  resp_var <- enquo(response)
  lme_mod <- lmer(as.formula(paste0(response, "~ start_y_c + (1|lake)")), data = dat)
  out <- summary(lme_mod)
  pred.mm <- ggpredict(lme_mod, terms = c("start_y_c"))
  
  plot <- ggplot(pred.mm) +
     geom_line(aes(x = x, y = predicted)) +
     geom_ribbon(aes(x = x, ymin = predicted - 1.96* std.error,
                     ymax = predicted + 1.96 * std.error),
                 fill = "lightgrey", alpha = 0.5) +
     geom_point(data = dat,
                aes_string(y = response,
                    x = "start_y_c",
                    color = "lake")) +
     theme_bw() +
     labs(title = title,
       caption = paste0(response, ". Line is predicted fixed effect from mixed model. Ribbon is 95% CI. Color points refere to lake ID"))
  print(plot)
  return(out)
}
```

* takes response variable as a quoted argument  

  * i.e., "ice_days" is the duration  
  
* Can also add the title as a quoted argument  

  * If no argument is supplied title is `NA`  
  
* returns the summary from the lmer model:  

  * formula = `response_variable ~ start_y_c + (1|lake)`  
  
* Also plots the estimated fixed effects  

### Duration

```{r message=FALSE, warning=FALSE}
mod_fun(full_data, "ice_days", "Ice Duration")
```
$\beta = -0.1976$; Decline of 0.1976 days per year (p < 0.001)
Random effects explain `r round(347.7 / (347.7+394.2),2)` % of the variation in the data

### Ice on

```{r message=FALSE, warning=FALSE}
mod_fun(full_data, "j_on_wy", "Ice On")
```

$\beta = 0.0968$; increase of 0.0968 days per year (p < 0.001)
Random effects explain `r round(270.8 / (270.8+238.1), 2)` % of the variation in the data

### Ice Off
```{r message=FALSE, warning=FALSE}
mod_fun(full_data, "j_off_wy", "Ice Off")
```
$\beta = -0.0940$; Decline of 0.0940 days per year (p < 0.001)
Random effects explain `r round(61.49 / (61.49+65.77), 2)` % of the variation in the data


## individual OLS

```{r message=FALSE, warning=FALSE}
ggplot(full_data, 
       aes(x = start_y_c,
                   y = ice_days,
                   color = lake)) +
     geom_point() +
  stat_smooth(method = "lm", se = FALSE) +
     theme_bw() +
  labs(title = "Duration")
```

#### Notice that Yellowstone is ~flat

```{r message=FALSE, warning=FALSE}
ggplot(full_data, 
       aes(x = start_y_c,
                   y = j_on_wy,
                   color = lake)) +
     geom_point() +
  stat_smooth(method = "lm", se = FALSE) +
     theme_bw() +
  labs(title = "ON")
```

#### Notice that Yellowstone is ~flat, also a few others maybe?

```{r message=FALSE, warning=FALSE}
ggplot(full_data, 
       aes(x = start_y_c,
                   y = j_off_wy,
                   color = lake)) +
     geom_point() +
  stat_smooth(method = "lm", se = FALSE) +
     theme_bw() +
  labs(title = "OFF")
```

#### Notice that Yellowstone + Baikal is ~flat