---
title: "non YSL mixed effects"
author: "Justin Pomeranz"
date: "8/22/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# libraries
library(tidyverse)
library(lme4)
library(lmerTest)
library(ggeffects)
```

# mixed effects for ice lake phenology
# Justin Pomeranz <jpomeranz@coloradomesa.edu>
# August 2023

### libraries

```{r}
library(tidyverse)
library(lme4)
library(lmerTest)
library(ggeffects)
```

```{r}
# Custom function from Oleksy
# custom function to caluclate julian-day of water year
hydro.day = function(x, start.month = 10L) {
  start.yr = year(x) - (month(x) < start.month)
  start.date = make_date(start.yr, start.month, 1L)
  as.integer(x - start.date + 1L)
}
```


### data
```{r message=FALSE, warning=FALSE}
## Non-Yellowstone lakes
non_ysl <- read_csv(here::here("scripts/other_phenology.txt"))

non_ysl <- non_ysl %>%
  # select necessary columns
  select(lake, lakecode, start_year, iceOn, iceOff, orig_duration) %>% 
  # filter out lakes: 
  filter(lakecode == "JK25"| # Lake Haukivesi
           lakecode == "JK02"| # Lake Kallavesi
           lakecode == "GW369"| # Lake Kallsjön
           lakecode == "JK03"| # Näsijärvi
           lakecode == "JK05"| # Päijänne
           lakecode == "JK40"| # Pielinen
           lakecode == "NG1" # Lake Baikal
  ) %>%
  # caluclate on and off julian dates for water year
  mutate(j_on_wy = hydro.day(iceOn),
         j_off_wy = hydro.day(iceOff)) %>%
  # filter out data only from 1927 to 2022
  filter(start_year >= 1927, start_year <=2022) %>%
  mutate(ice_days = j_off_wy - j_on_wy)

# yellowstone data
ysl_ice <- read.csv(here::here("scripts/YSLoff.csv")) %>%
  select(Year, IceOnDate, IceOffDate,) %>%
  mutate(lake = "yellowstone")

# modify data
ysl_ice <- ysl_ice %>%
  unite(IceOnDate, c(IceOnDate, Year), sep = "-", remove = FALSE) %>%
  unite(IceOffDate, c(IceOffDate, Year), sep = "-", remove = FALSE) %>%
  mutate(on_date = dmy(IceOnDate),
         off_date = dmy(IceOffDate)) %>%
  mutate(j_on_wy = hydro.day(on_date),
         j_off_wy = hydro.day(off_date),
         ice_days = j_off_wy - j_on_wy) %>%
  rename(start_year = Year) %>%
  select(lake, start_year, j_on_wy, j_off_wy, ice_days) 

# combine data sets
full_data <- bind_rows(non_ysl, ysl_ice) %>%
  mutate(start_y_c = start_year - mean(start_year, na.rm = TRUE))
```


## models

* year has been centered and = fixed  

* lake = random  

Model Function. 

```{r}
mod_fun <- function(dat, response, title = NA){
  resp_var <- enquo(response)
  lme_mod <- lmer(as.formula(paste0(response, "~ start_y_c + (1|lake)")), data = dat)
  out <- summary(lme_mod)
  pred.mm <- ggpredict(lme_mod, terms = c("start_y_c"))
  
  plot <- ggplot(pred.mm) +
     geom_line(aes(x = x, y = predicted)) +
     geom_ribbon(aes(x = x, ymin = predicted - 1.96* std.error,
                     ymax = predicted + 1.96 * std.error),
                 fill = "lightgrey", alpha = 0.5) +
     geom_point(data = dat,
                aes_string(y = response,
                    x = "start_y_c",
                    color = "lake")) +
     theme_bw() +
     labs(title = title,
       caption = paste0(response, ". Line is predicted fixed effect from mixed model. Ribbon is 95% CI. Color points refere to lake ID"))
  print(plot)
  return(out)
}
```

* takes response variable as a quoted argument  

  * i.e., "ice_days" is the duration  
  
* Can also add the title as a quoted argument  

  * If no argument is supplied title is `NA`  
  
* returns the summary from the lmer model:  

  * formula = `response_variable ~ start_y_c + (1|lake)`  
  
* Also plots the estimated fixed effects  

### Duration

```{r message=FALSE, warning=FALSE}
mod_fun(full_data, "ice_days", "Ice Duration")
```
$\beta = -0.1976$; Decline of 0.1976 days per year (p < 0.001)
Random effects explain `r round(347.7 / (347.7+394.2),2)` % of the variation in the data

### Ice on

```{r message=FALSE, warning=FALSE}
mod_fun(full_data, "j_on_wy", "Ice On")
```

$\beta = 0.0968$; increase of 0.0968 days per year (p < 0.001)
Random effects explain `r round(270.8 / (270.8+238.1), 2)` % of the variation in the data

### Ice Off
```{r message=FALSE, warning=FALSE}
mod_fun(full_data, "j_off_wy", "Ice Off")
```
$\beta = -0.0940$; Decline of 0.0940 days per year (p < 0.001)
Random effects explain `r round(61.49 / (61.49+65.77), 2)` % of the variation in the data


## individual OLS

```{r message=FALSE, warning=FALSE}
ggplot(full_data, 
       aes(x = start_y_c,
                   y = ice_days,
                   color = lake)) +
     geom_point() +
  stat_smooth(method = "lm", se = FALSE) +
     theme_bw() +
  labs(title = "Duration")
```

#### Notice that Yellowstone is ~flat

```{r message=FALSE, warning=FALSE}
ggplot(full_data, 
       aes(x = start_y_c,
                   y = j_on_wy,
                   color = lake)) +
     geom_point() +
  stat_smooth(method = "lm", se = FALSE) +
     theme_bw() +
  labs(title = "ON")
```

#### Notice that Yellowstone is ~flat, also a few others maybe?

```{r message=FALSE, warning=FALSE}
ggplot(full_data, 
       aes(x = start_y_c,
                   y = j_off_wy,
                   color = lake)) +
     geom_point() +
  stat_smooth(method = "lm", se = FALSE) +
     theme_bw() +
  labs(title = "OFF")
```

#### Notice that Yellowstone + Baikal is ~flat



# GAMS

```{r}
library(mgcv)
library(gratia)
mod0_iceOn <- gam(j_on_wy ~ s(start_year),
                        family=Gamma(link="log"),
                        data = full_data %>% 
                    filter(lake == "baikal"),
                        correlation = corCAR1(form = ~ start_year),
                        method = "REML")
summary(mod0_iceOn)
draw(mod0_iceOn)
appraise(mod0_iceOn)
```



```{r}
mod_fun2 <- function(data, gam_formula, lake_name){
  mod <- gam(gam_formula,
             family=Gamma(link="log"),
             data = data %>%
               filter(lake == lake_name),
             correlation = corCAR1(
               form = ~ start_year),
             method = "REML")
  mod_summary <- summary(mod)
  draw_plot <- draw(mod)
  app_plot <- appraise(mod)
  return(list(lake_name = lake_name,
              mod = mod,
              summary = mod_summary,
              draw_plot = draw_plot,
              appraise = app_plot))}
```

# test function

```{r}
mod_fun2(full_data, j_on_wy~s(start_year), "baikal")
mod_fun2(full_data, j_off_wy~s(start_year), "baikal")
mod_fun2(full_data, ice_days~s(start_year), "baikal")
mod_fun2(full_data, j_on_wy~s(start_year), "yellowstone")
```

```{r}
foo2 <- function(mydata, gam_formula, group_var) {
  group_var = enquo(group_var)

  mydata <- mydata %>% 
    group_by(!!group_var) %>% 
    nest() 

  mydata %>% 
    mutate(
      model = map(data,
                  ~(gam(
                    gam_formula,
                    family=Gamma(link="log"),
                    data = .,
                    correlation = corCAR1(
                      form = ~ start_year),
                    method = "REML")))) 
}
foo2(full_data, j_on_wy~s(start_year), lake) %>%
  mutate(tidied = map(model, tidy)) %>%
  unnest(tidied)
```

# testing foo 2

```{r}
library(broom)
foo2(full_data, j_on_wy~s(start_year), lake)
```


# list

For loop producing a list of the summary outputs for each lake. 

Run the loop for each response variable

### ice on
```{r}
lake_names <- full_data %>% pull(lake) %>% unique()

ice_on_list <- list()
for (lake in 1:length(lake_names)){
  lake_name = lake_names[lake]
  response_mod <- mod_fun2(full_data, j_on_wy~s(start_year), lake_name)
  ice_on_list[[lake]] <- response_mod
}
names(ice_on_list) <- lake_names
```

### ice off
```{r}
ice_off_list <- list()
for (lake in 1:length(lake_names)){
  lake_name = lake_names[lake]
  response_mod <- mod_fun2(full_data, j_off_wy~s(start_year), lake_name)
  ice_off_list[[lake]] <- response_mod
}
names(ice_off_list) <- lake_names
```



### duration
```{r}
ice_days_list <- list()
for (lake in 1:length(lake_names)){
  lake_name = lake_names[lake]
  response_mod <- mod_fun2(full_data, ice_days~s(start_year), lake_name)
  ice_days_list[[lake]] <- response_mod
}
names(ice_days_list) <- lake_names
```

# print out results of baikal as example

```{r}
list(ice_days_list$baikal,
     ice_off_list$baikal,
     ice_on_list$baikal)
```

## Yellowstone

```{r}
list(ice_days_list$yellowstone,
     ice_off_list$yellowstone,
     ice_on_list$yellowstone)
```

```{r}
ice_on_list
```

```{r}
ice_off_list
```


```{r}
ice_days_list
```

```{r}
ggplot(full_data,
       aes( x = start_year,
            y = ice_days,
            color = lake)) +
         geom_point() +
         stat_smooth(method = "gam") +
  facet_wrap(~lake)

ggplot(full_data,
       aes( x = start_year,
            y = j_on_wy,
            color = lake)) +
         geom_point() +
         stat_smooth(method = "gam") +
  facet_wrap(~lake)

ggplot(full_data,
       aes( x = start_year,
            y = j_off_wy,
            color = lake)) +
         geom_point() +
         stat_smooth(method = "gam") +
  facet_wrap(~lake)
```




## auto corr

```{r}
ACF <- acf(resid(ice_on_list$baikal$mod, type = "response"), plot = FALSE)
ACF <- setNames(data.frame(unclass(ACF)[c("acf", "lag")]), c("ACF","Lag"))
ggplot(ACF, aes(x = Lag, y = ACF)) +
  geom_hline(aes(yintercept = 0)) +
  geom_segment(mapping = aes(xend = Lag, yend = 0))
```

